---
title: "THE BIG PROJECT"
author: "Mohamed_Hassan"
date: "2025-10-19"
output: html_document
---
#Loading required libraries
```{r }
library(GEOquery)
gse_list <- unique(c("GSE48350","GSE5281","GSE15222","GSE11882")) 
GSE <- lapply(c(gse_list), function(x) getGEO(x)[[1]])
names(GSE) <- gse_list
```

##  Save lists of studies



```{r}
arraysP=GSE
dir.create("./Data", recursive = TRUE)
save(arraysP, file = "./Data/arrays.RData")
```

##Choose the right study
##good

```{r}
array = arraysP[["GSE48350"]]                                                  #
meta350=pData(array) 
sum(is.na(meta350$`gender:ch1`))
```

```{r}
array = arraysP[["GSE5281"]]
meta281=pData(array)
sum(is.na(meta281$`Sex:ch1`))
```
## 56


## not good 

```{r}
array = arraysP[["GSE15222"]]   
meta222=pData(array)
sum(is.na(meta222$`gender:ch1`))
```

##187

## not good 

```{r}
array = arraysP[["GSE11882"]]
meta882=pData(array)
sum(is.na(meta882))

```

##0 

##GSE11882 only control not good 


# Standarization of all ExpressionSets to make easier further analyses.
# We save three columns from phenotipic data:
## sex("male" or "female")
## brain_region(Options will be set to "HC","SFG","PCG" and "EC" in 03 script)
# Current categories are kept for a better understanding of data with
# data exploratory analysis.
## disease("case" or "control")

```{r}
library(Biobase)
array = arraysP[["GSE48350"]]

meta350 = pData(array)                  

dplot350 = data.frame(sex=tolower(meta350$`gender:ch1`),
                      brain_region=tolower(meta350$`brain region:ch1`),
                      group=tolower(meta350$title))

rownames(dplot350) = rownames(meta350)

dplot350$group = as.character(dplot350$group)

dplot350$group[grepl(pattern = "ad",x=dplot350$group)] <- "case"
dplot350$group[grepl(pattern = "indiv",x=dplot350$group)] <- "control"                     
colnames(dplot350)[3] = "disease"
dplot350 = dplot350[,c("sex","brain_region","disease")]
pData(array) = dplot350
array = array[complete.cases(exprs(array)),]
arraysP[["GSE48350"]] = array

save(dplot350, file = "./Data/dplot350.RData")    
dir.create("./Output", recursive = TRUE)
save(arraysP, file = "./Output/arrays_st.RData")
```

##data filtering and counting



```{r}
library(dplyr)

print(
  paste("male  samble count =",nrow(dplot350 %>% 
       filter(sex=="male"))))
print(
  paste("male control samble count =",nrow(dplot350 %>% 
       filter(sex=="male",disease == "control"))))
  
  dplot350 %>%
 filter(sex == "male", disease == "control") %>%
  count(brain_region)

```


```{r}
library(dplyr)

print(
  paste("male  samble count =",nrow(dplot350 %>% 
       filter(sex=="male"))))
print(
  paste("male case samble count =",nrow(dplot350 %>% 
       filter(sex=="male",disease == "case"))))
  
  dplot350 %>%
 filter(sex == "male", disease == "case") %>%
  count(brain_region)
```


```{r}
library(dplyr)

print(
  paste("female  samble count =",nrow(dplot350 %>% 
       filter(sex=="female"))))
print(
  paste("female control samble count =",nrow(dplot350 %>% 
       filter(sex=="female",disease == "control"))))
  
  dplot350 %>%
 filter(sex == "female", disease == "control") %>%
  count(brain_region)
```


```{r}
print(
  paste("female  samble count =",nrow(dplot350 %>% 
       filter(sex=="female"))))
print(
  paste("female case samble count =",nrow(dplot350 %>% 
       filter(sex=="female",disease == "case"))))
  
  dplot350 %>%
 filter(sex == "female", disease == "case") %>%
  count(brain_region)
```

# Here we normalize expression data when necessary,
# annotate probes, filter outliers, adapt brain region
# codes (phenotipic data) and select only samples from
# the selected brain regions ( "PCG","SFG","HC","EC" ).


```{r}

library(Biobase)
library(dplyr)
library(oligo)


```
# Function to pick median values when duplicated genes

```{r}
medianReps <- function(matriz){
  ID <- as.character(rownames(matriz))
  ID <- factor(ID, levels = unique(ID))
  df <- by(matriz, ID, function(x) apply(x, 2, stats::median))
  mat <- do.call("rbind", df)
  return(mat)
}
```

######### GSE48350 #########
# Preproccess expression matrix with log2 transform
# minimum value is set to 1 before transformation to get minimum equal to 0

```{r}
array <- arraysP[["GSE48350"]]
exprs(array) <- exprs(array) + (abs(min(exprs(array))) + 1)
exprs(array) <- log2(exprs(array))
```

# Formatting of brain region column and filter selected regions
```{r}
df <- pData(array)
df$brain_region <- as.character(df$brain_region)
df$brain_region[grepl("post-central gyrus",x = df$brain_region)] <- "PCG"
df$brain_region[grepl("postcentral gyrus",x = df$brain_region)] <- "PCG"
df$brain_region[grepl("superior frontal gyrus",x = df$brain_region)] <- "SFG"
df$brain_region[grepl("hippocampus",x = df$brain_region)] <- "HC"
df$brain_region[grepl("entorhinal cortex",x = df$brain_region)] <- "EC"
pData(array) <- df
array <- array[,df$brain_region %in% c("PCG","SFG","HC","EC")]

```


# Annotation of probes with gene symbol and Entrez ID
# Select median values when duplicated genes

```{r}
genedata <- fData(array)
colnames(genedata)
colnames(genedata)[colnames(genedata) == "Gene Symbol"] <- "Symbol"
colnames(genedata)[colnames(genedata) == "ENTREZ_GENE_ID"] <- "Entrez_Gene_ID"
colnames(genedata)
genedata <- genedata[,c("Symbol","Entrez_Gene_ID")]
genedata$Symbol <- sub(" ///.*","",genedata$Symbol)
genedata$Entrez_Gene_ID <- sub(" ///.*","",genedata$Entrez_Gene_ID)
exprs0 <- exprs(array)
rownames(exprs0) <- genedata$Symbol
exprs0 <- medianReps(exprs0)
exprs0 <- exprs0[rownames(exprs0) != "",]
genedata <- genedata[match(rownames(exprs0),genedata$Symbol),c("Symbol","Entrez_Gene_ID")]
rownames(genedata) <- genedata$Symbol
```
# Build the new preprocessed ExpressionSet and save it in list of arrays

```{r}
proc_array <- new("ExpressionSet",
                  exprs=exprs0,
                  annotation=annotation(array))
fData(proc_array) <- genedata
pData(proc_array) <- pData(array)

arraysP[["GSE48350"]] <- proc_array

```


#Differential expression analysis (DEA)
# with the following contrasts:
# contrast 1 -> femalecase - femalecontrol,
# contrast 2 -> malecase - malecontrol,
# contrast 3-> malecase - femalecase,
# contrast 4-> all case -all control
# DEA applied to ALL tissues,"PCG","SFG","HC","EC"

# load libraries

```{r}
library(Biobase)
library(limma)
library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)      # Database للجينات البشرية
library(ReactomePA)
library(enrichplot)
library(ggrepel)
library(ggplot2)
library(ggplot2)
library(ggrepel)
```

# Differential expression function
```{r}
diffexprs <- function(X,batch=F){
  sexdisease=vector("list",ncol(X))
  for(i in seq_along(sexdisease)) {
    sexdisease[[i]]=paste(pData(X)[,"sex"][i],pData(X)[,"disease"][i],sep="")
  }
  
  sexdisease=factor(unlist(sexdisease))
  
  # Consider batch effect or not(default)
  if (batch==T){
    batch=factor(pData(X)[,"batch"])
    design <- model.matrix(~ 0 + sexdisease + batch)
  } else {
    design <- model.matrix(~ 0 + sexdisease)
  }
  
  colnames(design)[1:4] <- levels(sexdisease)
  fit <- lmFit(X,design)
  cont.matrix <- makeContrasts(dif1=femalecase - femalecontrol,
                               dif2=malecase - malecontrol,
                               dif12=malecase -femalecase,
                               dif4=(femalecase + malecase)/2 - (femalecontrol + malecontrol)/2,
                               levels = design)
  fit2 <- contrasts.fit(fit,cont.matrix)
  fit2 <- eBayes(fit2)
  return(fit2)
}


wd <- "./Results/DEA/"
dir.create(wd, recursive = TRUE)

# Empty lists of results
ALLfit <- list()
HCfit<- list()
PCGfit <- list()
ECfit <- list()
SFGfit <- list()
for (study in names(arraysP)){
  array <- arraysP[[study]]
  # All tissues DEA
  fit <- diffexprs(array)
  ALLfit[[study]] <- fit
  # Write significant
  for (i in c(1:4)){
    cont <- topTable(fit,coef = i,adjust="BH",sort="none",n=Inf)
    cont <- cont[cont$adj.P.Val<0.05,]
    write.csv(cont,paste0(wd,study,"_ALL_cont",i,".csv"))
  }
  # HC differential expression
  if("HC" %in% pData(array)$brain_region) {
    filtered <- array[,pData(array)$brain_region == "HC"]
    fit <- diffexprs(filtered)
    HCfit[[study]] <- fit
    for (i in c(1:4)){
      cont <- topTable(fit,coef = i,adjust="BH",sort="none",n=Inf)
      cont <- cont[cont$adj.P.Val<0.05,]
      write.csv(cont,paste0(wd,study,"_HC_cont",i,".csv"))
    }
  }
  # PCG differential expression
  if("PCG" %in% pData(array)$brain_region) {
    filtered <- array[,pData(array)$brain_region == "PCG"]
    fit <- diffexprs(filtered)
    PCGfit[[study]] <- fit
    for (i in c(1:4)){
      cont <- topTable(fit,coef = i,adjust="BH",sort="none",n=Inf)
      cont <- cont[cont$adj.P.Val<0.05,]
      write.csv(cont,paste0(wd,study,"_PCG_cont",i,".csv"))
    }
  }
  # EC differential expression
  if("EC" %in% pData(array)$brain_region) {
    filtered <- array[,pData(array)$brain_region == "EC"]
    fit <- diffexprs(filtered)
    ECfit[[study]] <- fit
    for (i in c(1:4)){
      cont <- topTable(fit,coef = i,adjust="BH",sort="none",n=Inf)
      cont <- cont[cont$adj.P.Val<0.05,]
      write.csv(cont,paste0(wd,study,"_EC_cont",i,".csv"))
    }
  }
  #SFG differential expression
  if("SFG" %in% pData(array)$brain_region) {
    filtered <- array[,pData(array)$brain_region == "SFG"]
    fit <- diffexprs(filtered)
    SFGfit[[study]] <- fit
    for (i in c(1:4)){
      cont <- topTable(fit,coef = i,adjust="BH",sort="none",n=Inf)
      cont <- cont[cont$adj.P.Val<0.05,]
      write.csv(cont,paste0(wd,study,"_SFG_cont",i,".csv"))
    }
  }
}
# Save lists of "fit" objects
save(ALLfit, file = "./Output/ALLfit.RData")
save(HCfit, file = "./Output/HCfit.RData")
save(PCGfit, file = "./Output/PCGfit.RData")
save(ECfit, file = "./Output/ECfit.RData")
save(SFGfit, file = "./Output/SFGfit.RData")
```


```{r}

library(ggplot2)
library(ggrepel)

treasure_theme=theme(
  
  plot.title = element_text(hjust = 0.5,size=30,face = "bold"),
  
  axis.text.x = element_text( size = 15, angle = 45, hjust = .5, vjust = 0.5, face = "bold"),
  
  axis.text.y = element_text( size = 15, angle = 0, hjust = 1, vjust = 0, face = "bold"),
  
  axis.title.x = element_text( size = 20, angle = 0, hjust = .5, vjust = 0, face = "bold"),
  
  axis.title.y = element_text( size = 20, angle = 90, hjust = .5, vjust = .5, face = "bold"),
  
  #legend.title=element_text(size=20),
  
  legend.title=element_blank(), # remove legend title name
  
  legend.text = element_text(size=15,face="plain"),
  
  strip.text = element_text(size = 15,face="plain") ,
  
  legend.position="right",
  
  # for transparent background
  
  panel.background = element_rect(fill = "transparent"), # bg of the panel
  
  plot.background = element_rect(fill = "transparent", color = NA), # bg of the plot
  
  panel.grid.major = element_blank(), # get rid of major grid
  
  panel.grid.minor = element_blank(), # get rid of minor grid
  
  legend.background = element_rect(fill = "transparent"), # get rid of legend bg
  
  legend.box.background = element_rect(fill = "transparent"), # get rid of legend panel bg
  
  axis.line = element_line(colour = "black") # adding a black line for x and y axis
  
)
make_expr_function = function(name, expr_body, args = alist(x = )) {
 
  expr_body = substitute(expr_body)
  
  
  f = function() {}
  cat("✅ Step 1: Empty function created\n")
  
  
  formals(f) <- args
  cat("✅ Step 2: Arguments set to:\n")
  print(formals(f))
  
  
  body(f) = expr_body
  cat("✅ Step 3: Body set to:\n")
  print(body(f))
  
  assign(name, f, envir = .GlobalEnv)
  cat("✅ Step 4: Function assigned to name:", name, "\n")
  
  message(paste("🎉 Function", name, "created successfully!"))
}
```


#Functionmaker
#volcano_plot


```{r}
make_expr_function("volcano_plot",args = alist(data = , top_n = 50, fc_cutoff = 1, treasure_theme = NULL),
  expr_body = {
    
    # إنشاء عمود group
    data$group <- ifelse(data$logFC > fc_cutoff, "Up",
                         ifelse(data$logFC < -fc_cutoff, "Down", "Moderate"))
    
    # اختيار أعلى N جينات من كل نوع
    top_up <- data %>%
      filter(group == "Up") %>%
      arrange(desc(logFC)) %>%
      head(top_n)
    
    top_down <- data %>%
      filter(group == "Down") %>%
      arrange(logFC) %>%
      head(top_n)
    
    top_genes <- rbind(top_up, top_down)
    
    # رسم الفولكينو بلوت
    p <- ggplot(data, aes(x = logFC, y = -log10(adj.P.Val), color = group)) +
      geom_point(alpha = 0.8, size = 2) +
      scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Moderate" = "grey")) +
      geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed", color = "black") +
      geom_text_repel(
        data = top_genes,
        aes(label = Symbol),
        size = 3,
        max.overlaps = 20
      ) +
      labs(
        title = "Volcano Plot with Top Genes",
        x = "log2 Fold Change",
        y = "-log10(adj.P.Val)",
        color = ""
      )
    
    if (!is.null(treasure_theme)) p <- p + treasure_theme
    
    print(p)
  })
# volcano_plot(data = ALL_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```




#Functionmaker

#Enrishmint 



```{r}
make_expr_function(
  name = "enrichment_function",
  expr_body = {
    library(clusterProfiler)
    library(org.Hs.eg.db)
    library(ReactomePA)
    library(enrichplot)
    library(ggplot2)
    
    cat("🚀 Running enrichment pipeline on dataset...\n")
    
    # فلترة الجينات المهمة
    deg_data <- subset(x, adj.P.Val < 0.05)
    
    # تقسيم الجينات إلى up/down
    up_ids   <- deg_data$Entrez_Gene_ID[deg_data$logFC > 0]
    down_ids <- deg_data$Entrez_Gene_ID[deg_data$logFC < 0]
    
    # ---------------- GO Enrichment ----------------
    ego_up <- enrichGO(gene          = up_ids,
                       OrgDb         = org.Hs.eg.db,
                       ont           = "BP",
                       pAdjustMethod = "BH",
                       pvalueCutoff  = 0.05,
                       readable      = TRUE)
    
    ego_down <- enrichGO(gene          = down_ids,
                         OrgDb         = org.Hs.eg.db,
                         ont           = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff  = 0.05,
                         readable      = TRUE)
    
    # ---------------- KEGG Enrichment ----------------
    ekegg_up <- enrichKEGG(gene         = up_ids,
                           organism     = 'hsa',
                           pvalueCutoff = 0.05)
    
    ekegg_down <- enrichKEGG(gene         = down_ids,
                             organism     = 'hsa',
                             pvalueCutoff = 0.05)
    
    # ---------------- Reactome Enrichment ----------------
    ereact_up <- enrichPathway(gene         = up_ids,
                               organism     = "human",
                               pvalueCutoff = 0.05,
                               readable     = TRUE)
    
    ereact_down <- enrichPathway(gene         = down_ids,
                                 organism     = "human",
                                 pvalueCutoff = 0.05,
                                 readable     = TRUE)
    
    # ---------------- عرض النتائج ----------------
    cat("\n🧠 Top Reactome (Up):\n")
    print(head(ereact_up))
    cat("\n🧠 Top Reactome (Down):\n")
    print(head(ereact_down))
    
  # ---------------- Bar plots ----------------
print(barplot(ego_up, showCategory = 10, title = "GO BP - Upregulated"))
print(barplot(ego_down, showCategory = 10, title = "GO BP - Downregulated"))

print(barplot(ekegg_up, showCategory = 10, title = "KEGG - Upregulated"))
print(barplot(ekegg_down, showCategory = 10, title = "KEGG - Downregulated"))

print(barplot(ereact_up, showCategory = 10, title = "Reactome - Upregulated"))
print(barplot(ereact_down, showCategory = 10, title = "Reactome - Downregulated"))

# ---------------- Emapplots ----------------
if (nrow(as.data.frame(ereact_up)) > 1) {
  print(emapplot(pairwise_termsim(ereact_up), showCategory = 10) +
        ggtitle("Reactome Enrichment - Upregulated genes"))
}

if (nrow(as.data.frame(ereact_down)) > 1) {
  print(emapplot(pairwise_termsim(ereact_down), showCategory = 10) +
        ggtitle("Reactome Enrichment - Downregulated genes"))
}
    
    cat("✅ Enrichment analysis completed!\n")
    
    # رجّع النتائج كـ list لو حبيت تستخدمها
    list(
      GO_up = ego_up,
      GO_down = ego_down,
      KEGG_up = ekegg_up,
      KEGG_down = ekegg_down,
      Reactome_up = ereact_up,
      Reactome_down = ereact_down
    )
  },
  args = alist(x = )
)


#result= enrichment_function(x)



#frist datafram =head(ereact_up)
#second datafram=head(ereact_down)
```

# contrast 1 -> femalecase - femalecontrol,
#import dataset 


#*GSE48350_ALL_count1.csv=ALL_count1

```{r}
#ALL_cont1//volcano_plot
 volcano_plot(data = ALL_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```


```{r}
#ALL_cont1//enrichment
enrichment_function(ALL_cont1)
```


#*GSE48350_EC_count1.csv=EC_cont1

```{r}
#EC_cont1//volcano_plot

volcano_plot(data = EC_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```
#*GSE48350_HC_count1.csv=HC_cont1

```{r}
#HC_cont1//volcano_plot

volcano_plot(data = HC_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```
#*GSE48350_PCG_count1.csv=PCG_cont1

```{r}

#PCG_cont1//volcano_plot
volcano_plot(data = PCG_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)


```

#*GSE48350_SFG_count1.csv=SFG_cont1

```{r}
#SFG_cont1//volcano_plot

volcano_plot(data = SFG_cont1, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)

```



# contrast 2 -> malecase - malecontrol,
#import dataset 


#*GSE48350_ALL_count2.csv=ALL_count2

```{r}
#ALL_cont2//volcano_plot
 volcano_plot(data = ALL_cont2, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```


```{r}
#ALL_cont2//enrichment
enrichment_function(ALL_cont2)
```


#*GSE48350_EC_count2.csv=EC_cont2

```{r}
#EC_cont1//volcano_plot

volcano_plot(data = EC_cont2, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```

#*GSE48350_HC_count2.csv=HC_cont2

```{r}
#HC_cont2//volcano_plot

volcano_plot(data = HC_cont2, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```

#*GSE48350_PCG_count2.csv=PCG_cont2

```{r}
#PCG_cont2//volcano_plot
volcano_plot(data = PCG_cont2, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)

```

#*GSE48350_SFG_count2.csv=SFG_cont2

```{r}
#SFG_cont1//volcano_plot

volcano_plot(data = SFG_cont2, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)

```

# contrast 3-> malecase - femalecase,
#import dataset 

#*GSE48350_ALL_count3.csv

```{r}
#ALL_cont3//volcano_plot
 volcano_plot(data = ALL_cont3, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```


```{r}
#ALL_cont3//enrichment
enrichment_function(ALL_cont3)
```


#*GSE48350_EC_count3.csv=NULL
#*GSE48350_HC_count3.csv=NULL
#*GSE48350_PCG_count3.csv
#*GSE48350_SFG_count3.csv


```{r}


#PCG_cont3//volcano_plot
volcano_plot(data = PCG_cont3, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)



#SFG_cont3//volcano_plot

volcano_plot(data = SFG_cont3, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)


```

# contrast 4-> all case -all control

#import dataset 
#*GSE48350_ALL_count4.csv

```{r}
#ALL_cont4//volcano_plot
 volcano_plot(data = ALL_cont4, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)
```


```{r}
#ALL_cont4//enrichment
enrichment_function(ALL_cont4)
```


#*GSE48350_EC_count4.csv
#*GSE48350_HC_count4.csv
#*GSE48350_PCG_count4.csv
#*GSE48350_SFG_count4.csv


```{r}
#EC_cont4//volcano_plot

volcano_plot(data = EC_cont4, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)

#HC_cont4//volcano_plot

volcano_plot(data = HC_cont4, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)

#PCG_cont4//volcano_plot
volcano_plot(data = PCG_cont4, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)



#SFG_cont4//volcano_plot

volcano_plot(data = SFG_cont4, top_n = 50, fc_cutoff = 1, treasure_theme = treasure_theme)


```



##########################################################################




#pathfindR: An R Package for Enrichment Analysis Utilizing Active Subnetworks





#############################################################################






```{r}
install.packages("pak") # if you have not installed "pak"
pak::pkg_install("pathfindR")
```

```{r}
library(pathfindR)


```



```{r}
ALL_cont4_in=data.frame(
  Gene_symbol = ALL_cont4$Symbol,
  logFC = ALL_cont4$logFC,
  FDR_p = ALL_cont4$adj.P.Val)


output_ALL_cont4_ST <- run_pathfindR(ALL_cont4_in, gene_sets="Reactome",
                                     pin_name_path = "STRING",output_dir="./Results/Reactome_STRING")





```








```{r}
clustered_df_fuzzy_ST <- cluster_enriched_terms(output_ALL_cont4_ST, method="fuzzy")

# حفظ النتائج في ملف CSV
write.csv(clustered_df_fuzzy_ST, file = "./Cluster/clustered_df_fuzzy_ST.csv", row.names = FALSE)
```

#Heatmap

```{r}
# مكتبات
library(ggplot2)
library(reshape2)
library(dplyr)

# قراءة ملف CSV
df <- read.csv("./Cluster/clustered_df_fuzzy_ST.csv")

# اختيار أعلى 30 term حسب Fold_Enrichment
top_terms <- df %>%
  arrange(desc(Fold_Enrichment)) %>%
  head(50)

# تحويل البيانات لصيغة heatmap (Term_Description vs Cluster)
df_heat <- top_terms %>%
  select(Term_Description, Cluster, Fold_Enrichment) %>%
  dcast(Term_Description ~ Cluster, value.var = "Fold_Enrichment", fill = 0)

# تحويل للصيغة الطويلة للـ ggplot
df_long <- melt(df_heat, id.vars = "Term_Description")

# حفظ الرسم كـ PDF كبير وواضح
pdf("Fuzzy_Clustering_Heatmap.pdf", width = 14, height = 12)  # حجم كبير
ggplot(df_long, aes(x = variable, y = Term_Description, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  treasure_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  labs(
    x = "Cluster",
    y = "Term Description",
    fill = "Fold Enrichment",
    title = "Fuzzy Clustering Heatmap (Top 30 Terms)"
  )
dev.off()

# رسالة نجاح
cat("Heatmap تم حفظه كـ 'Fuzzy_Clustering_Heatmap.pdf'\n")


```


#(Network plot) ///////////   Interaction Diagram 

```{r}
# 1️⃣ تجهيز البيانات
input_processed <- input_processing(ALL_cont4_in)

# 2️⃣ اختيار المسارات الممثلة فقط
rep_terms <- subset(clustered_df_fuzzy_ST, Status == "Representative")

# 3️⃣ ترتيبهم حسب أقل p-value
rep_terms <- rep_terms[order(rep_terms$lowest_p), ]

# 4️⃣ اختيار أول 10 مسارات فقط
top10_terms <- head(rep_terms, 10)

# 5️⃣ رسم المسارات المختارة
gg_list <- visualize_terms(
  result_df = top10_terms,
  input_processed = input_processed,
  is_KEGG_result = FALSE,
  pin_name_path = "STRING"
)

# 6️⃣ إنشاء مجلد للحفظ (اختياري)
if (!dir.exists("Top10_Pathways")) dir.create("Top10_Pathways")

# 7️⃣ حفظ كل مسار كـ PDF منفصل
for (term_id in names(gg_list)) {
  ggplot2::ggsave(
    filename = paste0("Top10_Pathways/", term_id, ".pdf"),
    plot = gg_list[[term_id]],
    width = 10,
    height = 6
  )
  message("Saved: ", term_id)
}
```


```{r}
# 1️⃣ اختيار المسارات الممثلة فقط
rep_terms <- subset(clustered_df_fuzzy_ST, Status == "Representative")

# 2️⃣ استخراج الجينات من العمود الصحيح
rep_terms$Up_regulated <- as.character(rep_terms$Up_regulated)
rep_genes <- unlist(strsplit(rep_terms$Up_regulated, ";"))
rep_genes <- unique(trimws(rep_genes))

# 3️⃣ ربط الجينات بالتحليل التفاضلي الأصلي (DEGs)
bio_marker_candidates <- ALL_cont4_in %>%
  filter(Gene_symbol %in% rep_genes) %>%
  arrange(FDR_p, desc(abs(logFC)))

# 4️⃣ حفظ النتائج
if (!dir.exists("Biomarkers")) dir.create("Biomarkers")
write.csv(bio_marker_candidates, "./Biomarkers/biomarker_candidates.csv", row.names = FALSE)

cat("✅ تم استخراج الجينات المرشحة كـ biomarkers من المسارات الممثلة وحفظها في './Biomarkers/'\n")


```


```{r}
# 1️⃣ اختيار المسارات الممثلة فقط
rep_terms <- subset(clustered_df_fuzzy_ST, Status == "Representative")

# 2️⃣ استخراج الجينات من العمود الصحيح
rep_terms$Down_regulated <- as.character(rep_terms$Down_regulated)
rep_genes_dowen <- unlist(strsplit(rep_terms$Down_regulated, ";"))
rep_genes_dowen<- unique(trimws(rep_genes_dowen))

# 3️⃣ ربط الجينات بالتحليل التفاضلي الأصلي (DEGs)
bio_marker_candidates_dowen <- ALL_cont4_in %>%
  filter(Gene_symbol %in% rep_genes_dowen) %>%
  arrange(FDR_p, desc(abs(logFC)))

# 4️⃣ حفظ النتائج
if (!dir.exists("Biomarkers")) dir.create("Biomarkers")
write.csv(bio_marker_candidates_dowen, "./Biomarkers/biomarker_candidates_dowen.csv", row.names = FALSE)

cat("✅ تم استخراج الجينات المرشحة كـ biomarkers من المسارات الممثلة وحفظها في './Biomarkers/'\n")


```


```{r}
ALL_cont4_in_p=data.frame(
  Gene_symbol = ALL_cont4$Symbol,
  logFC = ALL_cont4$logFC,
  FDR_p = ALL_cont4$P.Value)


output_ALL_cont4_ST_p <- run_pathfindR(ALL_cont4_in_p, gene_sets="Reactome",
                                     pin_name_path = "STRING",output_dir="./Results_p/Reactome_STRING")
```



```{r}
genes_df <- data.frame(
  Gene_symbol = ALL_cont4_in_p$Gene_symbol,
  CHANGE = ALL_cont4_in_p$logFC,       # logFC للقيم
  P.Value = ALL_cont4_in_p$FDR_p     # optional لو مطلوب
)

# رسم heatmap لأكثر 10 مسارات دلالة
term_gene_heatmap(
  result_df = result_df,
  genes_df = genes_df,
  num_terms = 30,
  use_description = FALSE,
  low = "red",
  mid = "black",
  high = "green",
  legend_title = "logFC",
  sort_terms_by_p = TRUE
)
```


```{r}
# تحميل المكتبة (لو مش محملها)
library(pathfindR)

# نفترض output_df هي النتائج الناتجة من pathfindR
# الكلاستر باستخدام complete linkage
clustered_df_st_p <- cluster_enriched_terms(
  output_ALL_cont4_ST_p,
  clu_method = "complete",  # طريقة التجميع
  plot_hmap = TRUE,         # رسم heatmap للتشابه بين المصطلحات
  plot_dend = TRUE          # رسم dendrogram لتوضيح العلاقات
)

# عرض أول 6 نتائج مع cluster assignment
head(clustered_df_st_p)

```


```{r}
clustered_df_fuzzy_ST_p <- cluster_enriched_terms(output_ALL_cont4_ST_p, method="fuzzy")
```


```{r}
rep_terms <- subset(clustered_df_fuzzy_ST_p, Status == "Representative")

# 2️⃣ استخراج الجينات من العمود الصحيح
rep_terms$Down_regulated <- as.character(rep_terms$Down_regulated)
rep_genes_dowen <- unlist(strsplit(rep_terms$Down_regulated, ";"))
rep_genes_dowen<- unique(trimws(rep_genes_dowen))

# 3️⃣ ربط الجينات بالتحليل التفاضلي الأصلي (DEGs)
bio_marker_candidates_dowen <- ALL_cont4_in_p %>%
  filter(Gene_symbol %in% rep_genes_dowen) %>%
  arrange(FDR_p, desc(abs(logFC)))

# 4️⃣ حفظ النتائج
if (!dir.exists("Biomarkers_p")) dir.create("Biomarkers_p")
write.csv(bio_marker_candidates_dowen, "./Biomarkers_p/biomarker_candidates_dowen.csv", row.names = FALSE)

cat("✅ تم استخراج الجينات المرشحة كـ biomarkers من المسارات الممثلة وحفظها في './Biomarkers/'\n")

```


```{r}

# 1️⃣ اختيار المسارات الممثلة فقط
rep_terms <- subset(clustered_df_fuzzy_ST_p, Status == "Representative")

# 2️⃣ استخراج الجينات من العمود الصحيح
rep_terms$Up_regulated <- as.character(rep_terms$Up_regulated)
rep_genes <- unlist(strsplit(rep_terms$Up_regulated, ";"))
rep_genes <- unique(trimws(rep_genes))

# 3️⃣ ربط الجينات بالتحليل التفاضلي الأصلي (DEGs)
bio_marker_candidates_u <- ALL_cont4_in_p %>%
  filter(Gene_symbol %in% rep_genes) %>%
  arrange(FDR_p, desc(abs(logFC)))

# 4️⃣ حفظ النتائج
if (!dir.exists("Biomarkers_p")) dir.create("Biomarkers_p")
write.csv(bio_marker_candidates_u, "./Biomarkers_p/biomarker_candidates_u.csv", row.names = FALSE)

cat("✅ تم استخراج الجينات المرشحة كـ biomarkers من المسارات الممثلة وحفظها في './Biomarkers/'\n")

```





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
