---
title: "wgcnabrain"
author: "Mohamed_Hassan"
date: "2025-11-04"
output: html_document
---
```{r}
  devtools::install_github("kevinblighe/CorLevelPlot")

```

```{r}
# script to perform WGCNA
  

library(WGCNA)
library(DESeq2)
library(GEOquery)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra)

allowWGCNAThreads()          # allow multi-threading (optional)
```
# 1. Fetch Data

```{r #1. Fetch Data}
library(GEOquery)
gse_list <- unique(c("GSE48350")) 
GSE <- lapply(c(gse_list), function(x) getGEO(x)[[1]])
names(GSE) <- gse_list
arraysP=GSE

```


```{r #1. Fetch Data}
array = arraysP[["GSE48350"]]   
meta350=pData(array) 
phenoData <- meta350[,c(1,2,48,47)]
```


```{r #1.Fetch Data}

array = arraysP[["GSE48350"]]   
data=exprs(array) 
data[1:10,1:10]

class(array)
```

# prepare data
```{r #prepare data}



```

# 2. QC - outlier detection ------------------------------------------------
# detect outlier genes

```{r}
gsg <- goodSamplesGenes(t(data) )
summary(gsg)
gsg$allOK


```


```{r}
table(gsg$goodGenes)
table(gsg$goodSamples)
```


# detect outlier samples - hierarchical clustering - method 1

```{r}
htree <- hclust(dist(t(data)), method = "average")
plot(htree)
```

# pca - method 2 (another way)

```{r}
pca <- prcomp(t(data))
 pca.dat <- pca$x
 pca.var <- pca$sdev^2
 pca.var.percent <- round(pca.var/sum(pca.var)*100, digits = 2)
pca.dat <- as.data.frame(pca.dat)

 ggplot(pca.dat, aes(PC1, PC2)) +
  geom_point() +
  geom_text(label = rownames(pca.dat)) +
  labs(x = paste0('PC1: ', pca.var.percent[1], ' %'),
       y = paste0('PC2: ', pca.var.percent[2], ' %'))
 
```

### NOTE: If there are batch effects observed, correct for them before moving ahead
# exclude outlier samples

```{r}
samples.to.be.excluded <- c('GSM300303', 'GSM300272', 'GSM300172','GSM300171','GSM300305','GSM300320')
data.subset <- data[,!(colnames(data) %in% samples.to.be.excluded)]

```


# 3. Normalization ----------------------------------------------------------------------
# create a deseq2 dataset

# exclude outlier samples

```{r}
library(dplyr)

colData <- phenoData %>% 
  filter(!row.names(.) %in% samples.to.be.excluded)

```

# fixing column names in colData

```{r}
names(colData)
names(colData) <- gsub(':ch1', '', names(colData))
names(colData) <- gsub('\\s', '_', names(colData))
names(colData)
```


# making the rownames and column names identical

```{r}
all(rownames(colData) %in% colnames(data.subset))
all(rownames(colData) == colnames(data.subset))
```


# 4. Network Construction  ---------------------------------------------------
# Choose a set of soft-thresholding powers

```{r}
power <- c(c(1:10), seq(from = 12, to = 50, by = 2))

```

# Call the network topology analysis function

```{r}

norm.counts <- t(data.subset)

sft <- pickSoftThreshold(norm.counts,
                  powerVector = power,
                  networkType = "signed",
                  verbose = 5)


sft.data <- sft$fitIndices
```

# visualization to pick power

```{r}


a1 <- ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.6, color = 'red') +
  labs(x = 'Power', y = 'Scale free topology model fit, signed R^2') +
  theme_classic()


a2 <- ggplot(sft.data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Power', y = 'Mean Connectivity') +
  theme_classic()
  

grid.arrange(a1, a2, nrow = 2)

```
# convert matrix to numeric

```{r}

norm.counts[] <- sapply(norm.counts, as.numeric)

soft_power <- 18
temp_cor <- cor
cor <- WGCNA::cor

```

# memory estimate w.r.t blocksize
```{r}

bwnet <- blockwiseModules(norm.counts,
                 maxBlockSize = 14000,
                 TOMType = "signed",
                 power = soft_power,
                 mergeCutHeight = 0.25,
                 numericLabels = FALSE,
                 randomSeed = 1234,
                 verbose = 3)


cor <- temp_cor

```

# 5. Module Eigengenes ---------------------------------------------------------



```{r}
module_eigengenes <- bwnet$MEs

```

# Print out a preview

```{r}
head(module_eigengenes)
```




# get number of genes for each module



```{r}
table(bwnet$colors)
```

# Plot the dendrogram and the module colors before and after merging underneath
# grey module = all genes that doesn't fall into other modules were assigned to the grey module

```{r}
geneTree <- hclust(as.dist(1 - TOMsimilarityFromExpr(norm.counts, power = 18)), method = "average")

plotDendroAndColors(
  geneTree,
  bwnet$colors,
  "Merged modules",
  dendroLabels = FALSE,
  addGuide = TRUE,
  hang = 0.03
)


```


# 6A. Relate modules to traits --------------------------------------------------
# module trait associations



# create traits file - binarize categorical variables

```{r}
traits <- colData %>% 
  mutate(disease_state_bin = ifelse(grepl('AD', title), 1, 0)) %>% 
  select(5)


```


# binarize categorical variables

```{r}

colData$gender <- factor(colData$gender, levels = c("male", "female"))

gender.out <- binarizeCategoricalColumns(colData$gender,
                           includePairwise = FALSE,
                           includeLevelVsAll = TRUE,
                           minCount = 1)


traits <- cbind(traits, gender.out)


```

# Define numbers of genes and samples

```{r}
dim(module_eigengenes)
# [1] 50 10   ← 50 عينة × 10 modules

dim(traits)
# [1] 48 2    ← 48 عينة × 2 traits

```

```{r}
nSamples <- nrow(norm.counts)
nGenes <- ncol(norm.counts)

module.trait.corr <- cor(module_eigengenes, traits, use = 'p')
module.trait.corr.pvals <- corPvalueStudent(module.trait.corr, nSamples)
```


# visualize module-trait association as a heatmap

```{r}

heatmap.data <- merge(module_eigengenes, traits, by = 'row.names')

head(heatmap.data)

heatmap.data <- heatmap.data %>% 
  column_to_rownames(var = 'Row.names')




CorLevelPlot(heatmap.data,
             x = names(heatmap.data)[32:33],
             y = names(heatmap.data)[1:31],
             col = c("blue1", "skyblue", "white", "pink", "red"))


module.gene.mapping <- data.frame(Gene = names(bwnet$colors),
                                  Module = bwnet$colors)

grey.gen <- module.gene.mapping %>% 
  filter(Module == 'grey') %>% 
  pull(Gene)

head(grey.gen)

```


# prepare data
```{r #prepare data}

genedata <- fData(array)

# probes اللي عايز تجيب أساميهم
probe_ids <- c("117_at", "121_at", "1255_g_at", "1294_at", "1316_at", "1320_at")

# استخراج الجينات المقابلة من genedata
genes_from_probes <- genedata[rownames(genedata) %in% probe_ids, ]
# عرض أسماء الجينات فقط (العمود Symbol)
gene_symbols <- genes_from_probes$`Gene Symbol`

# عرض النتيجة
gene_symbols

```

# 6B. Intramodular analysis: Identifying driver genes ---------------



# Calculate the module membership and the associated p-values

# The module membership/intramodular connectivity is calculated as the correlation of the eigengene and the gene expression profile. 
# This quantifies the similarity of all genes on the array to every module.


```{r}

module.membership.measure <- cor(module_eigengenes, norm.counts, use = 'p')
module.membership.measure.pvals <- corPvalueStudent(module.membership.measure, nSamples)


module.membership.measure.pvals[1:10,1:10]


```

# Calculate the gene significance and associated p-values



# Using the gene significance you can identify genes that have a high significance for trait of interest 
# Using the module membership measures you can identify genes with high module membership in interesting modules.



```{r}
gene.signf.corr <- cor(norm.counts, traits$disease_state_bin, use = 'p')
gene.signf.corr.pvals <- corPvalueStudent(gene.signf.corr, nSamples)


gene.signf.corr.pvals %>% 
  as.data.frame() %>% 
  arrange(V1) %>% 
  head(25)

```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```
